Are we learning to build better models at the cost of better decisions?

### 2021-03-02 manuscript/manuscript.Rmd
:pen:

I illustrate the process of model selection by strictly proper scoring rules using two scenarios. In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig].  The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Gneiting2007,

@@

Each of these approaches swiftly decides that model 1 in our example is implausible, and settles for making decisions driven almost exclusively by model 2.  I illustrate this using a passive adaptive management strategy following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty. To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.  After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig]. As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone]. If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).

@@

Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.  The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig4b], while Model 2 leads to over-harvesting.  So why do our best approaches fail such a simple test case? It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved. I address these in turn.

@@

Is this result an artifact of the simple models considered here?  A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton Holt [@], Ricker [@], Shepard [@].  Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here. Simplicity facilitates intuition: Model 1 performs best because it correctly approximates the stock size which achieves maximum growth rate under the true model.  In contrast, each of those widely used recruitment relationships are symmetric, and their best-fitting parameter estimates will always have a maximum growth rate at a lower stock size. It is easy to point out that even this wider array of parameter and model values is too simple, since fisheries models in practice are much more complicated still (indeed, too complicated for the formal integration over model uncertainty considered here).  It is indeed true that either the forecasting or adaptive management example would work as expected if the set of candidate models included the true model, or at least a functional form that could better approximate the peak of the true model while also fitting better than model 2.  But we never have the true model in real systems.   If the candidate models are only caricatures, so too is the "true" model.  The candidate models considered here are no doubt much closer approximations to model 3 (all being single species, unstructured in age, stage, time or space, etc) than any model used in practice is to reality. Morever, a more accurate model is not required to reach optimal outcomes -- Model 1 already provides this by correctly approximating the key feature in this problem, even while it is wrong about much else.  How do proceed when our best methods fail in such a simple case?

@@

Is this result merely an artifact of differing objectives?  @Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.  However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.  Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1. To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.  This is the opposite recommendation of _adaptive_ management  @Dietze2018

Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world.  This is not an easy assertion to prove either way.  Future observations consistently fall outside the 95% confidence intervals predicted under Model 1.  Unfortunately, such innaccurate predictions are only too common in the real world.  Have any of those led to effective policy despite the innnaccuracy?  Have any been replaced with more accurate models that have also driven worse decisions? That is more difficult to answer.  What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.

>>>>

I illustrate the process of model selection by strictly proper scoring rules using two scenarios. In Scenario A (passive observation) the fish stock is unharvested and allowed to recover towards carrying capacity (as simulated under our "true" model, Model 3) while comparing the observed stock size in each subsequent time step to the distribution predicted under Model 1 and Model 2 respectively [Fig 1].  The mean, $\mu$ and variance, $\sigma$ of the forecast are compared against the true observation $x$ using a Proper scoring rule of @Gneiting2007,

@@

Each of these approaches swiftly decides that model 1 in our example is implausible, and settles for making decisions driven almost exclusively by model 2.  I illustrate this using a passive adaptive management strategy following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty. To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.  After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 4]. As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone]. If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).

@@

Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.  The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  So why do our best approaches fail such a simple test case? It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved. I address these in turn.

@@

Is this result an artifact of the simple models considered here?  A more realistic approach would consider uncertainty over model parameters, as well as models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980].  Such complexity would not be sufficient to resolve the paradox of model uncertainty illustrated here. Simplicity facilitates intuition: Model 1 performs best because it correctly approximates the stock size which achieves maximum growth rate under the true model.  In contrast, each of those widely used recruitment relationships are symmetric, and their best-fitting parameter estimates will always have a maximum growth rate at a lower stock size. It is easy to point out that even this wider array of parameter and model values is too simple, since fisheries models in practice are much more complicated still (indeed, too complicated for the formal integration over model uncertainty considered here).  It is indeed true that either the forecasting or adaptive management example would work as expected if the set of candidate models included the true model, or at least a functional form that could better approximate the peak of the true model while also fitting better than model 2.  But we never have the true model in real systems.   If the candidate models are only caricatures, so too is the "true" model.  The candidate models considered here are no doubt much closer approximations to model 3 (all being single species, unstructured in age, stage, time or space, etc) than any model used in practice is to reality. Morever, a more accurate model is not required to reach optimal outcomes -- Model 1 already provides this by correctly approximating the key feature in this problem, even while it is wrong about much else.  How do proceed when our best methods fail in such a simple case?

@@

Is this result merely an artifact of differing objectives?  @Gneiting2007's proper scoring rules are designed to select models which make the best forecast, and Model 2 really is the best choice for that task.  However, the adaptive management approach explicitly seeks to maximize expected utility, integrating over the uncertainty in the model, and yet it evolves away from Model 1 which would achieve that goal and settles on Model 2.  Note that this issue is not resolved by updating the learning step to reflect only the immediate reward -- it is in fact easy to see that so-called greedy algorithms would take the largest harvest possible and never reach the sustainable yield achieved by lower harvest rates of Model 1. To select Model 1 over alternatives like Model 2 which make better predictions but lead to worse outcomes requires a more long-term approach that forgoes iterative updating of probabilities as new data becomes available.  Note that this is the opposite recommendation of _adaptive_ management [@Walters1978] or _iterative_ forecasting [@Dietze2018].

Perhaps Model 1 is a unicorn, a mythically simple model that gives optimal results but has no analogue in the real world.  This is not an easy assertion to prove either way.  Future observations consistently fall outside the 95% confidence intervals predicted under Model 1.  Unfortunately, such innaccurate predictions are only too common in the real world.  Have any of those led to effective policy despite the innaccuracy?  Have any been replaced with more accurate models that have also driven worse decisions? That is more difficult to answer.  What this example demonstrates is that if such unicorn models exist, our current paradigm will surely filter them out eventually, to be replaced with worse outcomes.
